\thetitle

### 2008-06-12 figures/basic_behaviors.tex
Moved all tables & figures into separate files.

\documentclass[twocolumn,final]{svjour3}

\newcommand{\thetitle}{Linear Representation of Network Traffic}
\newcommand{\theauthors}{
  Stefan~Karpinski \and
  Elizabeth~M.~Belding \and
  Kevin~C.~Almeroth \and
  John~R.~Gilbert
}

%\usepackage{type1cm}
%\usepackage[pdftex]{graphicx}
\usepackage[labelfont=bf,small]{caption}
\usepackage[font=small,labelfont=bf,position=top,nearskip=0em]{subfig}
\usepackage{cite,amsmath,amssymb,rotating,multirow,bigstrut,url,wrapfig,relsize,paralist,array}
\usepackage[hyperfigures,bookmarks,bookmarksopen,bookmarksnumbered,colorlinks,linkcolor=black,citecolor=black,filecolor=blue,menucolor=black,pagecolor=blue,frenchlinks=true,pdftitle={\thetitle}]{hyperref}
\usepackage{mathtools}

\title{\thetitle}
\author{\theauthors}
\institute{
  S.~Karpinski \and
  E.~Belding \and
  K.~Almeroth \and
  J.~Gilbert \\
  Department of Computer Science \\
  University of California, Santa Barbara (USA) \\
  Email: \texttt{\{sgk,ebelding,almeroth,gilbert\}@cs.ucsb.edu}
}

\input{definitions}

\graphicspath{{graphics/}}

\begin{document}
\maketitle

\begin{abstract}
We propose a representation of wireless workload patterns as large, sparse matrices and provide a method for stochastically generating experimental workload from a given matrix. The essential property of the algebraic representation is that the summation of vectors naturally yields a faithful description of the aggregate behavior of the corresponding flows. This deceptively simple property allows us to express many common concepts from traffic modeling succinctly in terms of a few linear transformations. The algebraic representation has many benefits: 1)~it makes the meaning of generally understood but vague concepts, such as ``uniform behavior,'' mathematically precise and unambiguous; 2)~it allows us to see clearly, through the lens of linear algebra, the implications of common modeling assumptions; 3)~the implementation of traffic models becomes unprecedentedly simple and orthogonal, requiring only a handful of high-level matrix operations, which can be freely composed; 4)~the vast body of algebraic theory and highly optimized numerical linear software may immediately be applied to traffic modeling. We use the paired differential simulation methodology introduced by the authors in previous work to experimentally demonstrate that the general matrix model accurately reproduces realistic network performance~\cite{Karpinski07:realism,Karpinski07:cbr-failure}. We use the same experimental methodology to explore the implications of various assumptions and simplifications that are commonly made in traffic modeling.
%We demonstrate experimentally that the general matrix model accurately reproduces network performance metrics, using the differential simulation methodology introduced in~\cite{Karpinski07:realism} and applied in~\cite{Karpinski07:cbr-failure} to common uniform traffic models. A range of common modeling approaches are implemented in the general matrix model and evaluated experimentally.
%
%We demonstrate how linear algebra can be effectively used to represent and perform computations with traffic behaviors in wireless networks. Moreover, we map common concepts and techniques from traffic modeling into this new algebraic domain, showing how they can be expressed with mathematical precision and succinctness, while simultaneously being made clearer and more general. We demonstrate experimentally that the general matrix model accurately reproduces network performance metrics, using the differential simulation methodology introduced in~\cite{Karpinski07:realism} and applied in~\cite{Karpinski07:cbr-failure} to discredit the common constant bit-rate model. A range of common modeling approaches are implemented in the general matrix model and evaluated experimentally. We conclude with future directions for modeling research based on the new algebraic framework.
\end{abstract}

% The model of flow behavior implicit in non-negative matrix factoriation technique is that flow behavior is a composite of some limited set of basic behaviors---and that fractions of each flow ``belong'' to each behavior. This is not an \latin{a priori} unreasonable hypothesis. Contrast that with the assumptions underlying uniform and marginal simplifications of traffic behavior, which can be seen to be intuitively unreasonable without ever conducting and actual experiment or calculation.

\keywords{
  traffic modeling \and
  traffic analysis \and
  workload generation \and
  wireless networks \and
  wireless simulation \and
  linear algebra \and
  non-negative matrix factorization
}

\section{Introduction}\label{sec:intro}

\newfootnote{\flownote}{We use the common definition of a \textit{flow} as a sequence of packets sharing the same  ``5-tuple'': \caps{IP} protocol type, source and destination nodes, and \caps{TCP/UDP} port numbers.}

\newfootnote{\marginalnote}{The statistical term ``marginal'' refers to the margins of actuarial tables formerly used for statistical computations. The rows and columns of the table represent possible values of two properties. Each entry in a table contains a count of the number of events falling into the joint category for that row and column. The margins contain sums of the rows and columns, thus giving the unconditional distributions of each property.}

More than twenty years of research in analysis and modeling of network traffic have yielded tremendous advances in our understanding of the complex behaviors that emerge from the interaction of millions of humans and computers on the Internet and in local-area networks (\caps{LAN}s).
Both traffic analysis and realistic workload generation, however, remain active areas of research with many unanswered questions.
One of the major challenges of modeling whole-network traffic is the interdependence of packet, flow\flownote and node behaviors.
Three types of simplifying assumptions are commonly applied in both analysis and generation:
\begin{enumerate}
\item \textbf{Uniform:} that traffic properties are uniformly distributed. For example, assuming that  all nodes in a network have equal probability of being the source or destination of each flow.
Another common example is the assumption that all packet sizes and inter-packet intervals are the same; this is commonly known as the ``constant bit-rate'' (\caps{CBR}) model.
\item \textbf{Marginal:} that traffic properties share a single, common \emph{marginal}\marginalnote distribution across the entire network, and individual values are drawn independently and unconditionally from this shared distribution.
For example, assuming that a single distribution of packet sizes describes all flows in a network, as compared to equipping each flow with its own specific distribution of sizes.
\item \textbf{Conditional:} that traffic behaves marginally within equivalence classes defined by certain conditions.
The assumption, for example, that all flows emanating from a common source node share the same packet size distribution is a conditional model, conditioned on the source nodes.
\end{enumerate}
We would be arguing against a straw man if we implied that researchers who apply these assumptions actually believe them to be true.
They clearly are not:
different nodes initiate and receive drastically different numbers of packets and flows;
flows have drastically different packet size distributions.
Simply considering intuitive examples of common behavior illustrates this amply:
BitTorrent users \latin{vs.} occasional email checkers;
downloading a large file \latin{vs.} typing in an \caps{SSH} session.

% In short, the reason is that these assumptions drastically simplify both traffic analysis and workload generation.
% If we ignore the interdependencies that make analysis difficult, the difficulties simply evaporate. 
% This immediately begs the question of whether these assumptions are benign.
% Does assuming uniformity or marginality of network behavior distort the properties and metrics we are trying to analyze and replicate?
% This paper answers that question: we cannot stick our heads in the sand and pretend that network behaviors are uniform or marginal without severely distorting crucial network performance metrics.

% TODO: is this paragraph necessary?
% The second reason why uniformity and marginality assumptions are so commonly used is that there are no generally accepted alternatives for representing or analyzing traffic.
% The difficulty lies in the high-dimensional and stochastic nature of network behaviors.
% Suppose, for example, that we assume that packet sizes and inter-packet intervals are drawn independently from some given distribution for each flow.
% In the terminology introduced above, this is a flow-conditional packet behavior model.
% Prior research has indicated that, given accurate per-flow distributions, and keeping all other aspects of network behavior fixed, this simplification does not distort experimental performance metrics~\cite{Karpinski07:realism}.
% In this model, however, each flow must have its own specific, realistic pair of distributions for packet size and inter-packet intervals.
% Moreover, the ``distribution of distributions'' across the entire network must simultaneously be realistic.
% Modeling such distributions of distributions realistically for entire networks is currently an unsolved problem; we begin to address it here.

If no one actually believes that assumptions of uniformity and marginality accurately describe network behavior, then why are these assumptions so common?
There are two reasons.
The first is that these assumptions drastically simplify both traffic analysis and workload generation, making them tractable problems.
% If we ignore the interdependencies that make analysis difficult, the difficulties simply evaporate. 
% This immediately begs the question of whether these assumptions are benign.
% Does assuming uniformity or marginality of network behavior distort the properties and metrics we are trying to understand?
% This paper answers that question.
The second is that there are no generally accepted alternatives for representing and simplifying traffic patterns that allow effective analysis and generation without assuming uniformity or marginality.
% Representing complex multi-level network behaviors in a way that does not make these simplifying assumptions but allows effective traffic analysis and workload generation is currently an unsolved problem;
% we begin to address it here.

% This paper answers that question: we cannot stick our heads in the sand and pretend that network behaviors are uniform or marginal without severely distorting crucial network performance metrics.

Fundamentally, uniform, marginal and conditional models are applied so that network traffic behavior can be represented using a manageable collection of empirical or analytical statistical distributions.
Reduction in representation size and complexity makes analysis and generation feasible.
From this perspective, three questions immediately present themselves:
\begin{enumerate}
\item Do these assumptions distort the metrics we are trying to analyze, replicate, and predict?
\item Can we represent ``unreduced'' network behavior such that analysis and generation remain tractable?
\item Are there better ways to simplify and reduce the representation of network behavior?
\end{enumerate}
This paper answers all three questions. We begin, however by answering the second problem and  
using our solution to address the other two questions.
% transforming the unreduced representation of traffic into it's uniform, marginal or conditional counterparts.

To represent unreduced network behavior practically, we propose the concept of a \emph{linear representation} of network traffic.
A linear representation of traffic maps each flow to a vector, such that the following linearity condition is satisfied:
\begin{quote}
The sum of the representations of two flows is a vector that represents the aggregate behavior of the two flows.
\end{quote}
The complete behavior of a network is expressed as a matrix of such vectors, having one row per flow.
Linear representations express full network behavior, without imposing assumptions of uniformity or marginality, yet still allow effective traffic analysis and workload generation.
We present a specific linear representation, called the general matrix model (\caps{GMM}), and demonstrate experimentally that this model generates wireless workload that accurately reproduces the performance characteristics of real wireless network traffic.

\newfootnote{\ranknote}{For uniform and marginal assumptions, the multiplication matrix has rank 1  since all rows and columns are identical. In the conditional case, the rank of the matrix is the number of conditioning classes.}

To address the the other two questions, we begin with a simple observation.
In linear representations of traffic, uniform, marginal and conditional assumptions are assertions of equality between the rows and columns of the traffic matrix.
Moreover, real traffic matrices can be transformed into uniform, marginal and conditional forms through very simple matrix multiplications.
These transformations, however, are extremely destructive: the matrices are degenerate with very low rank, so multiplication destroys almost all information in the original traffic matrix.
We demonstrate experimentally, that this loss of information is not benign:
uniform and marginal traffic models severely distorting crucial network performance metrics.

\input{figures/trace_vs_uniform}

Once traffic is represented linearly, alternatives to assuming uniformity or marginality become apparent.
We argue that matrix factorization, which is nondestructive and does not assume uniformity or marginality, is far superior to multiplication for this problem.
To demonstrate this approach, we use non-negative matrix factorization (\caps{NMF})~\cite{Lee01} to find a low-dimensional approximation of the packet size and inter-packet interval matrices of a real traffic trace.
This factorization extracts a small set of ``basic behaviors'' from the traffic matrix and simultaneously explains each flow's observed behavior as a mixture of these basic behaviors.
Although the factorization is computed without examining port numbers, basic behaviors nevertheless correspond to intuitive expectations for protocols.
We find basic behaviors corresponding to typical network activities, including file transfer, typing via \caps{SSH}, web surfing, ping traffic, peer-to-peer and others.

% TODO: update roadmap paragraph.

The rest of the paper is organized as follows. In \Section{motivation} we discuss motivation and related work. The notion of linear representation and the general matrix model are presented in \Section{general-matrix-model}. In Sections~\ref{sec:common-properties}~and~\ref{sec:modeling-concepts} we explore how common network properties and modeling concepts can be expressed in the the \caps{GMM} framework. Our experimental and analytical methodology for evaluating traffic models is presented in \Section{methodology}, while the results of our experiments are explained and analyzed in \Section{results}. The ramifications of these results are discussed in \Section{discussion}. Finally, in \Section{conclusions}, we conclude with a discussion of how this research may be applied to current wireless studies.

\section{Motivation \& Related Work}
\label{sec:related-work}
\label{sec:motivation}

\newfootnote{\QoSnote}{This is violated by some quality of service (QoS) schemes. However, we can simply add QoS metadata---such as traffic classes or urgency flags---to our models of user behavior and the rest of our arguments remain valid. The network is still disinterested in the exact content of the data being transported; only the QoS metadata is relevant.}

The interaction of network users and applications with the lower layers of the networking stack is characterized by where, when, how much, and to whom data is transmitted. The joint pattern of traffic generation and mobility through time and space completely determines the effect of network usage on the lower levels of the stack. This is due to the data-agnostic nature of the protocol stack: by design, \caps{IP} networks treat all data in the same manner~\cite{Clark88}.\QoSnote The credibility of conclusions derived from simulation or experimental deployment depends crucially on our confidence that the models used to generate traffic in experiments are sufficiently realistic. \Figure{trace-vs-uniform} illustrates through examples how drastically different network traffic appears under assumptions of uniformity at different levels of behavior. Marginality assumptions are not as visually striking, but as we will demonstrate, they distort the performance characteristics of networks almost as much, and sometimes more, than uniform models.

While this work is not inherently specific to wireless networks, since the modeling techniques developed are equally applicable to traffic in wired networks, the results are of special importance to wireless research.
Realistic traffic models have a drastic impact on experimental results in wireless networks~\cite{Karpinski07:realism,Karpinski07:cbr-failure}.
This is primarily because wireless networks are locally resource constrained.
Compare this with the typical situation in wired \caps{LAN}s, where gigabit Ethernet and modern routers  can easily handle all but the most intense levels of traffic.
If a wired network operator expects larger volumes of traffic on a conditional basis, the solution is to upgrade the hardware, not to change the network protocols.
In wireless networks, on the other hand, the physical medium cannot be upgraded; protocols must be improved to make better usage of the medium.
Being able to perform experiments with realistic workload is a vital part of the ongoing process of improving wireless protocols.
Without realistic workload models for experiments, performance comparisons may be misleading or even completely wrong~\cite{Karpinski07:cbr-failure}.

Paxson and Floyd observed~\cite{Paxson95} that the interaction between endpoint behavior and the network conditions is inherently \textit{closed-loop} in the sense that the emergent behavior is governed by a closed feedback loop.
In the case of most non-\caps{TCP} applications, this feedback does not occur: an open-loop model suffices.
Hern\'andez-Campos' dissertation~\cite{Hernandez06:dissertation} addresses in great depth how to model the feedback between the network and typical \caps{TCP} applications.
He shows how to abstract an application's behavior from a \caps{TCP} trace and replay it in a manner that accurately reflects how typical applications react in response to different network conditions.
This research has subsequently been turned into a \caps{TCP} flow generation tool called \tmix~\cite{Weigle06}.
The acronym is somewhat misleading: {\tmix} does not provide models for application behaviors or for mixtures of applications in networks.
Rather, it provides the ability to reproduce accurate closed-loop dynamics for a single \caps{TCP} flow \emph{given} an application behavior model as input.
Our research complements this perfectly: we provide the means to analyze traffic and generate realistic whole-network application mixtures which can then serve as the necessary inputs to {\tmix}.

\newfootnote{\harpoonnote}{Even this is unclear: generated traffic is only compared with trace traffic using the very same metrics that are explicitly specified as part of the model. Unsurprisingly, the distributions of sampled metrics closely match the distributions those metrics were sampled from. Realism is not double-checked using other statistical metrics or actual performance metrics.}

The two most prominent general traffic generation frameworks are Harpoon and \caps{D-ITG}. Harpoon~\cite{Sommers04} uses a traffic trace for self-training, and can subsequently generate synthetic traffic with certain statistical properties based on the original trace.
The properties reproduced are the empirical distributions of the following: ``file size, inter-connection time, source and destination IP ranges, [and] number of active sessions.''
Harpoon exemplifies the uniform and marginal modeling approaches.
Source and destination nodes are treated marginally: empirically derived marginal frequencies of sources and destinations are used to choose endpoints independently for each flow.
Flow sizes (\latin{i.e.} file sizes) are likewise treated marginally: each flow's total byte  count is sampled from a marginal distribution of flow sizes.
\caps{UDP} flows are modeled as being constant bit-rate;
\caps{TCP} flows are all treated as file transfers, making them effectively uniform with bit-rate determined by \caps{TCP} dynamics.
Inter-connection times are sampled from a single empirically derived marginal distribution.
While Harpoon may be sufficiently realistic for the intended purpose of generating Internet backbone traffic,\harpoonnote for generating local-area traffic in wireless experiments, it is not.
Previous research has shown that uniform packet behavior models used in Harpoon drastically distort vital performance metrics at all levels of the protocol stack~\cite{Karpinski07:cbr-failure}.
This paper shows, moreover, that marginal models, like those used in Harpoon, also severely misrepresent performance metrics in wireless networks.

Like Harpoon, \caps{D-ITG}~\cite{Avallone04} exemplifies uniform and marginal modeling assumptions. However, unlike Harpoon, \caps{D-ITG} does not derive models from real traffic.
Packet behavior can be modeled using standard statistical distributions (``constant, uniform, normal, cauchy, pareto, exponential, etc.'').
Guidance is not provided, however, for choosing among these distributions, choosing parameter values, or selecting a mixture of distributions across a network.
Flow and node behaviors are not specified at all, but rather left to the user's discretion.
The focus of this project is primarily on the engineering challenge of generating very large volumes of synthetic traffic and injecting it into networks in a distributed fashion.
While \caps{D-ITG} provides the mechanism for generating large volumes of traffic, it does not provide any real guidance as to what traffic to generate.

\FHC~\cite{Hernandez06:spatio-temporal-modeling} have defined the state of the art in parametric modeling of wireless traffic patterns.
They use rigorous statistical methods to fit parametric models to a number of properties of wireless traffic.
Flow arrivals fit a non-stationary Poisson model;
flow inter-arrival times fit a Lognormal distribution;
the number of flows per session fits a BiPareto model;
flow sizes also fit a BiPareto model.
The statistical analysis fitting these parametric models is thorough and convincing.
The models, however, are marginal:
a single statistical model applies to all flows and nodes; packet behavior is not considered.
Each node has a different number of flows assigned to it, but otherwise nodes are identical. 
Moreover, all flows, regardless of source node, are statistically identical:
flow sizes are drawn from a single marginal distribution, and nothing else distinguishes them.
Yet, intuitively it is clear that different nodes in real networks have drastically different tendencies for flow size, flow rates, and packet behavior.
Such interdependent behaviors cannot be captured using marginal modeling.
In this paper, we demonstrate that this inability is not harmless:
treating all flows and nodes statistically identically severely distorts crucial wireless performance metrics in experiments.
We also provide sorely needed methods of representing and analyzing traffic patterns without making uniform or marginal assumptions about network behavior.

This work uses the technique known as \emph{paired differential simulation} to evaluate the realism of modified or synthetically generated wireless workloads.
This technique was proposed and subsequently refined by Karpinski~\latin{et~al.}~\cite{Karpinski07:realism,Karpinski07:cbr-failure}.
They introduce the notion of \emph{sufficient realism}, and propose paired differential simulation as an experimental methodology based on the standard technique of pairing control and test subjects.
We describe the technique in \Section{methodology}.
In the first paper~\cite{Karpinski07:realism}, they explore various ways of modifying trace traffic patterns, as a form of sensitivity analysis to determine which aspects of network behavior are essential and must be preserved in realistic models, and which aspects can be discarded as inessential.
Of particular interest is their conclusion that detailed time-series packet behavior for flows is inessential:
it suffices, for each flow, to have accurate unconditional distributions of packet sizes and inter-packet intervals;
sufficiently realistic flow behavior can be reconstructed from these distributions by repeated independent sampling.
In the second paper~\cite{Karpinski07:cbr-failure}, Karpinski~\latin{et~al.} explore the impact of a broad variety of uniformity assumptions at the packet, flow and node levels of behavior.
They find that all uniformity assumptions are detrimental, inducing significant and often extreme misrepresentation of a broad variety of wireless performance metrics at every level of the protocol stack.
Moreover, they show, using \caps{OLSR} and \caps{AODV}, that uniform traffic models can completely invert the relative performance of network protocols.

\section{Linear Representation of Traffic}
\label{sec:representations}

The traffic workload pattern in a network is a collection of \caps{IP} packets between hosts in the network sent at certain times.
It is standard in network analysis, however, to aggregate sequences of packets sharing the same ``5-tuple'':
\caps{IP} protocol, source and destination nodes, and source and destination \caps{TCP/UDP} port numbers.
Such a sequence of packets is called a \textit{flow}.
The traffic pattern of an entire network is simply the collected behaviors of all flows occurring in the network.

The behavior of a flow as it affects the network is characterized by the following properties:
its \caps{IP} protocol type (\caps{TCP}, \caps{UDP}, \caps{ICMP}, etc.);
its source and destination nodes;
its start time;
and the specific sizes packets of and intervals between their transmission.
We begin our exposition of traffic representation by formalizing these properties mathematically as sets:
\begin{tabbing}
\hspace{1em}\=\bullet\hspace{0.5em}\=\hspace{5em}\=\hspace{4em}\=$\in\Nodes\,$\=\kill
\>\bullet\>types:\>$\type$\>$\in\Types$\>$= \set{1,\dots,255} \subset \N$ \\
\>\bullet\>nodes:\>$\src,\dst$\>$\in\Nodes$\>$= \set{1,\dots,n} \subset \N$ \\
\>\bullet\>times:\>$\start$\>$\in\Times$\>$= \R$ \\
\>\bullet\>sizes:\>$\size$\>$\in\Sizes$\>$= \set{0,\dots,\text{\caps{MTU}}} \subset \N$ \\
\>\bullet\>intervals:\>$\ival$\>$\in\Ivals$\>$ = \R^+ \union \set{\infty}$.
\end{tabbing}
There are 255 possible \caps{IP} types (0 is reserved for \caps{IPv6}).
The number $n$ is simply the total number of nodes in the network being modeled.
\caps{MTU} is the maximum transfer unit of the network (typically around 1500 bytes).
We formalize the packet behavior of a flow as a pair of infinite sequences of packet sizes and inter-packet intervals:
\begin{align}
\sizes &= \seq{\size_1,\size_2,\dots} \in \SizeSeq \\
\ivals &= \seq{\ival_1,\ival_2,\dots} \in \IvalSeq.
\end{align}
Since flows contain only a finite number of packets, there exists a number of packets, $\pkts \in \N$, for each flow. Since flows are finite, we require that the sequences $\sizes$ and $\ivals$ satisfy the following requirements:
\begin{align}
&\forall\: k \le \pkts: \quad \size_k > 0       \eqn{positive-packets}  \\
&\forall\: k >   \pkts: \quad \size_k = 0       \eqn{zero-packets}      \\
&\forall\: k <   \pkts: \quad \ival_k < \infty  \eqn{finite-intervals}  \\
&\forall\: k \ge \pkts: \quad \ival_k = \infty. \eqn{infinite-intervals}
\end{align}
Packet sizes are required to be positive because we are modeling application layer data transfer:
an empty data transmission request is a non-event.
The send time of the $k$th packet can be expressed as
\begin{align}
\time_k = \start + \sum_{i=1}^{k-1} \ival_i.
\end{align}
This is guaranteed to be finite if and only if $k \le \pkts$.
Now we are ready to formalize the space of flows:
\begin{align}
  \Flows =
    \Types    \cross
    \Nodes^2  \cross
    \Times    \cross
    \N        \cross
    \SizeSeq  \cross
    \IvalSeq.
\end{align}
We write a generic element of $\Flows$ as
\begin{align}
  \flow = \tuple{
    \type,
    \src,
    \dst,
    \start,
    \pkts,
    \sizes,
    \ivals
  } \in \Flows.
\end{align}
Let $2^\Flows$ denote the power set of $\Flows$, and $\Traffic \subseteq 2^\Flows$ the set of all finite sets of flows:
\begin{align}
  \Traffic = \set{X \in 2^\Flows : \card{X} < \infty}.
\end{align}
A traffic sample is simply an element of $\Traffic$. We will refer to elements of $\Traffic$ interchangeably as ``traffic samples'' or ``traffic instances.''

Having chosen a mathematical formalization of network traffic, we may now define the notion of a linear representation of network traffic.

\begin{definition}\defn{linear-representation}
A \defterm{linear representation} of network traffic is a pair, $\tuple{V,\psi}$, where $V$ is a vector space and $\psi$ is a function, $\psi: \Traffic \to V$, such that
\begin{align}\eqn{linearity}
\forall~T \in \Traffic\!:\quad
  \psi(T) &= \sum_{\mathclap{\flow \in T}}{\psi(\flow)}.
\end{align}
\end{definition}

\noindent
In plain words, every finite collection of flows is represented by a vector that is the sum of the elements representing the individual flows in the collection.
% Representing traffic using vectors allows us to analyze, model, and understand traffic patterns in terms of the thoroughly studied and well understood subject of linear algebra.
The careful reader will notice that \emph{any} function $\Flows \to V$ uniquely determines a linear representation:
\Equation{linearity} uniquely extends the function to all of $\Traffic$.
Representing flows by arbitrary vectors, however, is not particularly useful or interesting.
We focus instead on meaningful representations, where collections of flows are represented by vectors that naturally describe the aggregate behavior in some manner.
In the next section, we clarify these concepts through examples.

\subsection{Examples of Linear Representations}
\sec{linear-representation-examples}

\newfootnote{\pbdnote}{There are many schemes to correct the discrepancy in bytes when $\pkts$ does not divide $\mathsym{bytes}$ evenly. The simplest is to ignore it; we leave leave more complex schemes to the reader.}

Our first example is the packets-bytes-duration (\caps{PBD}) representation.
The representation is $\pbd: \Traffic \to \R^3$, mapping
$\flow \mapsto \tuple{p,b,d} \in \R^3$, where
\begin{align}
p &= \pkts, &
b &= \sum_{i=1}^{\mathclap{p}}{\size_i}, &
d &= \sum_{i=1}^{\mathclap{p-1}}{\ival_i}.
\end{align}
This representation encapsulates the information necessary to reproduce a \caps{CBR} version of each flow:
packets have payloads of $\floor{\fracx{b}{p}}$ bytes, and are sent every $\fracx{d}{p}$ seconds.\pbdnote
Given two flows with behavior vectors $\vec{x}_1$ and $\vec{x}_2$, their sum as vectors, 
\begin{align}
\vec{x}_1+\vec{x}_2=\tuple{p_1+p_2,b_1+b_2,d_1+d_2},
\end{align}
provides an appropriate description of the aggregate behavior of the two flows together.
Similarly, their average, \latin{i.e.} $\frac{1}{2}(\vec{x}_1+\vec{x}_2)$, has the average number of packets, bytes, and duration, which is precisely what the ``average behavior'' of the two flows should intuitively be.
In contrast, a ``packets-payload-interval'' representation, expressing the \caps{CBR} parameters directly, does not have this property:
the sum of average payloads is not the overall average of payloads;
likewise for intervals.
In the \caps{PBD} representation, the linearity property leads to natural descriptions of aggregate behavior through vector addition, whereas in the packets-payload-interval representation, vector addition is ``unnatural.''

The \caps{PBD} representation is highly lossy: it discards most of the information detailing the behavior of each flow.
At the other extreme, we can give a ``perfect'' representation of each flow, that allows us to completely reconstruct each flow from its representation.
With care, the sum of representations of flows can still naturally describe their aggregate behavior.
Let the representation be $\Phi: \Traffic \to \FlowVec$, where
\begin{align}
\FlowVec = \R^{255} \cross \R^n \cross \R^n \cross L^2[\R] \cross L^2[\R].
\end{align}
% Here $\R^X$ is the vector space of functions $f: X \to \R$, which is isomorphic to the standard Euclidean vector space, $\R^\card{X}$.
Here $L^2[\R]$ is the vector space of real functions, $\R^\R=\set{f: \R \to \R}$, with the usual norm~\cite{Rudin87}.
% \begin{align}
% \norm{f}_2 = \Big[ \int_{-\infty}^{\infty} \abs{f}^2 dx \Big]^{\frac{1}{2}}.
% \end{align}
We define
\begin{align}
\Phi(\flow) = \tuple{\e_\type,\e_\src,\e_\dst,f,g},
\end{align}
where $\e_k$ is the $k$th standard basis vector, and $f,g \in L^2[\R]$ are functions defined by
\begin{align}
f(t) &= \sum_{k=1}^{\pkts}
  \begin{cases}
    1 & \text{if $t \ge \time_k$} \\
    0 & \text{if $t <   \time_k$},
  \end{cases} \\
g(t) &= \sum_{k=1}^{\pkts}
  \begin{cases}
    \size_k & \text{if $t \ge \time_k$} \\
    0       & \text{if $t <   \time_k$}.
  \end{cases}
\end{align}
Thus, $f$ and $g$ are both integer-valued, monotonically increasing step functions, giving the cumulative number of packets and total bytes sent, respectively, at time $t$.
From the representation, $\Phi(\flow)$, of a single flow, we can readily recover its type, source, destination, and the transmission times and sizes of packets.

Why is $\Phi$ natural in the same sense that the \caps{PBD} representation is natural?
Suppose we have a traffic sample, $T\in\Traffic$.
The first component of $\Phi(T)$ is a 255-dimensional vector, whose coordinates are a histogram of the number of times each flow type occurs in $T$.
Similarly, the second and third components of $\Phi(T)$ are $n$-dimensional histograms of the number of times each node occurs as source and destination.
The last components are step functions, like $f$ and $g$.
Vector addition in $L^2[\R]$ dictates that these functions give the total number of cumulative packets and bytes sent, respectively, over all flows in $T$.
Thus, summation of vectors under the representation $\Phi$ yields a natural and useful description of the aggregate behavior of flows.

Consider, for contrast, if we had represented flow type simply by $\type \in \N$ instead of $\e_\type \in \R^{255}$.
The natural numbers containing $\type$ can readily be embedded into $\R^1$, which is a vector space.
However, the addition of vectors is highly unnatural.
The sum of protocol numbers is meaningless:
two \caps{TCP} flows plus five \caps{ICMP} flows does not equal a single \caps{UDP} flow, despite the fact that $2\cdot6+5\cdot1=17$.
The summation of vectors in the representation $\Phi$, on the other hand is $5\,\e_1+2\,\e_6$, indicating the frequency with which each type occurred.

Both of the example representations discussed here can be applied in practice.
The \caps{PBD} representation naturally allows us to compute aggregate and average \caps{CBR} behaviors of collections of flows.
With a suitable machine encoding of step-functions, $\Phi$ can be implemented and used in practice.
It provides complete information about individual flows and highly detailed (though not complete) information about collections of flows.
Representations that preserve even more information about traffic instances can readily be constructed.
We turn now, however, in a different direction, and introduce a specific linear representation that strikes a practical balance in fidelity between $\pbd$ and $\Phi$.

\section{The General Matrix Model}
\sec{general-matrix-model}

\input{tables/gmm_properties}

The \caps{PBD} representation of flows uses 3 dimensions to encode the behavior of each flow, while $\Phi$ uses $255+2n$ dimensions, plus the continuum of dimensions of two Banach spaces (\latin{i.e.} the two copies of $L^2[\R]$).
In this section, we define a representation, $\phi$, that splits the difference and uses a large but finite number of dimensions to represent each flow.
Once the vector representation of each flow is defined, we may construct matrices of representation vectors, with a row for each flow in a traffic sample.
We call the matrix representation of network traffic the \emph{general matrix model}~(\caps{GMM}).
Having expressed network behavior in terms of matrices, we can very simply and efficiently compute a broad variety of network behavior properties using standard matrix operations.
Furthermore, we can express uniformity and marginality assumptions about behavior as equality constraints between the rows and columns of the traffic matrix.
Moreover, we can transform unconstrained traffic instances into uniform, marginal, or conditional approximations via right or left multiplications by easily computed matrices.

The general strategy for defining $\phi$ is to divide the descriptive properties of each flow into bins---%
one per value for properties that are discrete already and require no compression---%
and map values in each bin to a standard basis element of $\R^d$, where $d$ is the number of bins.
In this scheme, representation vectors are effectively histograms, with coordinate values in each dimension counting items falling into the corresponding bin.
% The histograms for packet sizes and inter-packet intervals count packets, while other histograms count flows.
\Table{quantization} lists the properties that are used to define $\phi$, together with the quantization and dequantization functions.
Each quantization function takes a possible property value as input and maps it to an index value in $\set{1,\dots,d}$, where $d$ is the number of bins, \latin{i.e.} dimension, for the property in question.
% The index assigned determines the histogram bin of the input property.

The histogram-vector approach produces some representations of properties that are not immediately intuitive, and may seem to use an excessive number of representation dimensions.
However, this approach automatically satisfies the ``naturalness'' criterion for linear representations.
That is, the representation, $\phi(T)$, naturally describes the aggregate behavior of flows in $T$.
Perhaps even more importantly, arbitrary linear combinations of vectors can be meaningfully interpreted as describing some realizable, hypothetical traffic instance.
In \Section{traffic-generation} we will describe how to generate a flow for any sum of representation vectors.
% This is how we generate workload from transformed traffic matrices.

>>>>

%!TEX root = ../paper.tex
